---
title: "Copynumber PCAs"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE/')
```

In this script I create different PCAs for different subsets of the HGDP summary dataset. The focus is the `copynumber` of every transposon for each individual.
```{r}
library(tidyverse)
library("ggpubr")
HGDP <- read_delim("/Volumes/Temp1/rpianezza/TE/summary-HGDP/USEME_HGDP_complete_reflib6.2_mq10_batchinfo_cutoff0.01.txt")
names(HGDP) <- c("ID","pop","sex","country","type","familyname","length","reads","copynumber","batch")

SGDP <- read_tsv("/Volumes/Temp2/rpianezza/SGDP/summary/USEME_SGDP_cutoff") %>% dplyr::rename(ID=biosample)

HGDP_pcr_free_samples <- read_tsv("/Volumes/Temp1/rpianezza/investigation/HGDP-no-PCR/HGDP-only-pcr-free-samples.tsv", col_names = "ID")
HGDP_pcr_free <- HGDP %>% filter(ID %in% HGDP_pcr_free_samples$ID)
HGDP_pcr <- HGDP %>% filter(!(ID %in% HGDP_pcr_free_samples$ID))

SGDP_pcr_free_samples <- read_tsv("/Volumes/Temp1/rpianezza/SGDP/ric-documentation/SGDP-no-PCR.tsv")
SGDP_pcr_free <- SGDP %>% filter(ID %in% SGDP_pcr_free_samples$ID)
```

# Function for PCA plotting

First, I create a function that takes a subset of the dataset (`data`) and a string as `title` for the plot title and gives back the complete PCA plot. Note that `ellipses` are shown in all the plots, to remove them you should modify the function here.
```{r}
PCA_copynumber <- function(raw_data, title){
data <- filter(raw_data, type=="te")
m <- filter(data, sex=='male')
f <- filter(data, sex=='female')
len <- length(unique(data$familyname))
males <- length(unique(m$ID))
females <- length(unique(f$ID))

f_matrix <- matrix(as.vector(f$copynumber),nrow=females,ncol=len,byrow=T)
f_fram <- data.frame(f_matrix)
names(f_fram) <- unique(f$familyname)
f_fram <- f_fram %>% select_if(negate(function(col) sd(col)==0))
f_matrixcont <- matrix(as.vector(f$country),nrow=females,ncol=len,byrow=T)
f_framcont <- data.frame(f_matrixcont)
f_contcol<-c(f_framcont$X1)

m_matrix <- matrix(as.vector(m$copynumber),nrow=males,ncol=len,byrow=T)
m_fram <- data.frame(m_matrix)
names(m_fram)<-unique(m$familyname)
m_fram <- m_fram %>% select_if(negate(function(col) sd(col)==0))
m_matrixcont <- matrix(as.vector(m$country),nrow=males,ncol=len,byrow=T)
m_framcont <- data.frame(m_matrixcont)
m_contcol<-c(m_framcont$X1)
  
fHGDP.pca <- prcomp(f_fram, center = TRUE, scale = TRUE)
f_var <- summary(fHGDP.pca)$importance[2,1]

fHGDP.pca <- prcomp(f_fram, center = TRUE, scale = TRUE)
f_var1 <- summary(fHGDP.pca)$importance[2,1]
f_var2 <- summary(fHGDP.pca)$importance[2,2]

f_PCA <- ggplot(data=f_fram, aes(x=fHGDP.pca$x[,1], y=fHGDP.pca$x[,2], color = f_contcol)) + 
geom_point() + 
ggtitle("Females") + 
xlab(paste0("PC1 (", round(f_var1,3)*100,"%)")) + 
ylab(paste0("PC2 (",round(f_var2,3)*100,"%)")) + 
theme(plot.title = element_text(size = 8, hjust = 0.5)) #+ 
guides(col = guide_colourbar(title = ""))

mHGDP.pca <- prcomp(m_fram, center = TRUE, scale = TRUE)
m_var1 <- summary(mHGDP.pca)$importance[2,1]
m_var2 <- summary(mHGDP.pca)$importance[2,2]

m_PCA <- ggplot(data=m_fram, aes(x=mHGDP.pca$x[,1], y=mHGDP.pca$x[,2], color = m_contcol)) +
geom_point() +
ggtitle("Males") +
xlab(paste0("PC1 (", round(m_var1,3)*100,"%)")) +
ylab(paste0("PC2 (",round(m_var2,3)*100,"%)")) +
theme(plot.title = element_text(size = 8, hjust = 0.5))

figure <- ggarrange(f_PCA, m_PCA, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))
annotate_figure(figure, top = text_grob(title, color = "black", size = 20), fig.lab = "")
}
```

```{r}
PCA_copynumber_SGDP <- function(raw_data, title){
  
color_palette <- c("Africa" = "#F8766D", "America" = "#C49A00", "Central Asia and Siberia" = "#53B400", "East Asia" = "#00C094", "West Eurasia" = "#00B6EB", "South Asia" = "#A58AFF", "Oceania" = "#FB61D7")
    
data <- filter(raw_data, type=="te")
m <- filter(data, sex=='male')
f <- filter(data, sex=='female')
len <- length(unique(data$familyname))
males <- length(unique(m$ID))
females <- length(unique(f$ID))

f_matrix <- matrix(as.vector(f$copynumber),nrow=females,ncol=len,byrow=T)
f_fram <- data.frame(f_matrix)
names(f_fram) <- unique(f$familyname)
f_fram <- f_fram %>% select_if(negate(function(col) sd(col)==0))
f_matrixcont <- matrix(as.vector(f$country),nrow=females,ncol=len,byrow=T)
f_framcont <- data.frame(f_matrixcont)
f_contcol<-c(f_framcont$X1)

m_matrix <- matrix(as.vector(m$copynumber),nrow=males,ncol=len,byrow=T)
m_fram <- data.frame(m_matrix)
names(m_fram)<-unique(m$familyname)
m_fram <- m_fram %>% select_if(negate(function(col) sd(col)==0))
m_matrixcont <- matrix(as.vector(m$country),nrow=males,ncol=len,byrow=T)
m_framcont <- data.frame(m_matrixcont)
m_contcol<-c(m_framcont$X1)
  
fHGDP.pca <- prcomp(f_fram, center = TRUE, scale = TRUE)
f_var <- summary(fHGDP.pca)$importance[2,1]

fHGDP.pca <- prcomp(f_fram, center = TRUE, scale = TRUE)
f_var1 <- summary(fHGDP.pca)$importance[2,1]
f_var2 <- summary(fHGDP.pca)$importance[2,2]

f_PCA <- ggplot(data=f_fram, aes(x=fHGDP.pca$x[,1], y=fHGDP.pca$x[,2], color = f_contcol)) + 
geom_point() + 
ggtitle("Females") + 
xlab(paste0("PC1 (", round(f_var1,3)*100,"%)")) + 
ylab(paste0("PC2 (",round(f_var2,3)*100,"%)")) + 
theme(plot.title = element_text(size = 8, hjust = 0.5))+ 
guides(col = guide_colourbar(title = ""))+ 
scale_color_manual(values = color_palette)

mHGDP.pca <- prcomp(m_fram, center = TRUE, scale = TRUE)
m_var1 <- summary(mHGDP.pca)$importance[2,1]
m_var2 <- summary(mHGDP.pca)$importance[2,2]

m_PCA <- ggplot(data=m_fram, aes(x=mHGDP.pca$x[,1], y=mHGDP.pca$x[,2], color = m_contcol)) +
geom_point() +
ggtitle("Males") +
xlab(paste0("PC1 (", round(m_var1,3)*100,"%)")) +
ylab(paste0("PC2 (",round(m_var2,3)*100,"%)")) +
theme(plot.title = element_text(size = 8, hjust = 0.5))+ 
scale_color_manual(values = color_palette)

figure <- ggarrange(f_PCA, m_PCA, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))
annotate_figure(figure, top = text_grob(title, color = "black", size = 20), fig.lab = "")
}
```

# HGDP

## PCA

```{r}
PCA_copynumber(HGDP, "HGDP - All TEs")
PCA_copynumber(HGDP_pcr_free, "HGDP - All TEs - PCR free samples")
PCA_copynumber(HGDP_pcr, "HGDP - All TEs - PCR samples")
```


## PCA for the most variable TEs

```{r}
variableTEs <- filter(HGDP_pcr_free, type == "te") %>% group_by(familyname) %>% summarise(variance = var(copynumber)) %>% filter(variance>0.5) %>% select(familyname) %>% pull()

HGDP_var <- filter(HGDP, familyname %in% variableTEs)
HGDP_var_pcrfree <- filter(HGDP_pcr_free, familyname %in% variableTEs)
HGDP_var_pcr <- filter(HGDP_pcr, familyname %in% variableTEs)

PCA_copynumber(HGDP_var, "HGDP - Most variable TEs")
PCA_copynumber(HGDP_var_pcrfree, "HGDP - Most variable TEs - PCR free samples")
PCA_copynumber(HGDP_var_pcr, "HGDP - Most variable TEs - PCR samples")
```

## PCA for unbiased GC content TEs

```{r}
GC <- read_tsv("/Volumes/Temp1/rpianezza/GC-content/gc-content.tsv") %>% filter(GC_content > 45 & GC_content < 55) %>% select(familyname) %>% pull()

HGDP_GC <- filter(HGDP, familyname %in% GC)
HGDP_GC_pcrfree <- filter(HGDP_pcr_free, familyname %in% GC)
HGDP_GC_pcr <- filter(HGDP_pcr, familyname %in% GC)

PCA_copynumber(HGDP_GC, "HGDP - GC TE content 45-55%")
PCA_copynumber(HGDP_GC_pcrfree, "HGDP - GC TE content 45-55% - PCR free samples")
PCA_copynumber(HGDP_GC_pcr, "HGDP - GC TE content 45-55% - PCR samples")
```


## PCA excluding biased samples (PCR samples + GC biased samples PCR free)

```{r}
a_HGDP <- read_tsv("/Volumes/Temp1/rpianezza/PCA-copynumber-all-analysis/a_HGDP.tsv")
HGDP_nobiased_samples <- filter(a_HGDP, (a > (-0.5)) & (a<0.5)) %>% select(ID) %>% pull()
HGDP_nobiasedID <- filter(HGDP, ID %in% HGDP_nobiased_samples)
HGDP_nobiasedID_pcrfree <- filter(HGDP_pcr_free, ID %in% HGDP_nobiased_samples)
HGDP_nobiasedID_pcrfree_varTE <- filter(HGDP_nobiasedID_pcrfree, familyname %in% variableTEs)

#PCA_copynumber(HGDP_nobiasedID, "HGDP - GC biased samples excluded")
PCA_copynumber(HGDP_nobiasedID_pcrfree, "HGDP - GC biased samples excluded")
PCA_copynumber(HGDP_nobiasedID_pcrfree_varTE, "HGDP - GC biased samples excluded - Variable TEs")
```

# SGDP

## PCA

```{r}
PCA_copynumber_SGDP(SGDP, "SGDP - All TEs")
PCA_copynumber_SGDP(SGDP_pcr_free, "SGDP - All TEs - PCR free samples")
```


## PCA for the most variable TEs

```{r}
variableTEs <- filter(SGDP_pcr_free, type == "te") %>% group_by(familyname) %>% summarise(variance = var(copynumber)) %>% filter(variance>0.5) %>% select(familyname) %>% pull()

SGDP_var <- filter(SGDP, familyname %in% variableTEs)
SGDP_var_pcrfree <- filter(SGDP_pcr_free, familyname %in% variableTEs)

PCA_copynumber_SGDP(SGDP_var, "SGDP - Most variable TEs")
PCA_copynumber_SGDP(SGDP_var_pcrfree, "SGDP - Most variable TEs - PCR free samples")
```


## PCA for unbiased GC content TEs

```{r}
GC <- read_tsv("/Volumes/Temp1/rpianezza/GC-content/gc-content.tsv") %>% filter(GC_content > 45 & GC_content < 55) %>% select(familyname) %>% pull()

SGDP_GC <- filter(SGDP, familyname %in% GC)
SGDP_GC_pcrfree <- filter(SGDP_pcr_free, familyname %in% GC)

PCA_copynumber_SGDP(SGDP_GC, "SGDP - GC TE content 45-55%")
PCA_copynumber_SGDP(SGDP_GC_pcrfree, "SGDP - GC TE content 45-55% - PCR free samples")
```


## PCA excluding biased samples

```{r}
a_SGDP <- read_tsv("/Volumes/Temp1/rpianezza/copynumberPCA-all-analysis/a_SGDP.tsv")
SGDP_nobiased_samples <- filter(a_SGDP, (a > (-0.5)) & (a<0.5)) %>% select(ID) %>% pull()
SGDP_nobiasedID <- filter(SGDP, ID %in% SGDP_nobiased_samples)
SGDP_nobiasedID_pcrfree <- filter(SGDP_pcr_free, ID %in% SGDP_nobiased_samples)
SGDP_nobiasedID_pcrfree_varTE <- filter(SGDP_nobiasedID_pcrfree, familyname %in% variableTEs)

PCA_copynumber_SGDP(SGDP_nobiasedID_pcrfree, "SGDP - GC biased samples excluded")
PCA_copynumber_SGDP(SGDP_nobiasedID_pcrfree_varTE, "SGDP - GC biased samples excluded - Variable TEs")
```

