---
title: "PCA for different subsets of the dataset"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE/')
```

In this script I create different PCAs for different subsets of the HGDP summary dataset. The focus is the `copynumber` of every transposon for each individual.
```{r}
library(tidyverse)
library("ggpubr")
HGDPcutoff <- read_delim("/Volumes/Temp1/rpianezza/TE/summary-HGDP/USEME_HGDP_complete_reflib6.2_mq10_batchinfo_cutoff0.01.txt")
names(HGDPcutoff) <- c("ID","pop","sex","country","type","familyname","length","reads","copynumber","batch")

HGDPcutoff <- HGDPcutoff %>% mutate(country = recode(country, "Oceania_(SGDP),Oceania"="Oceania"))
```
# Function for PCA plotting

First, I create a function that takes a subset of the dataset (`data`) and a string as `title` for the plot title and gives back the complete PCA plot. Note that `ellipses` are shown in all the plots, to remove them you should modify the function here.
```{r}
PCA <- function(data, title){
  m <- filter(data, sex=='male')
  f <- filter(data, sex=='female')
  len <- length(unique(data$familyname))
  males <- length(unique(m$ID))
  females <- length(unique(f$ID))

  f_matrix<-matrix(as.vector(f$copynumber),nrow=females,ncol=len,byrow=T)
  f_fram<-data.frame(f_matrix)

  names(f_fram)<-unique(f$familyname)
  f_matrixcont<-matrix(as.vector(f$country),nrow=females,ncol=len,byrow=T)
  f_framcont<-data.frame(f_matrixcont)
  f_contcol<-c(f_framcont$X1)

  fHGDP.pca <- prcomp(f_fram, center = TRUE, scale = TRUE)
  
  m_matrix<-matrix(as.vector(m$copynumber),nrow=males,ncol=len,byrow=T)
  m_fram<-data.frame(m_matrix)

  names(m_fram)<-unique(m$familyname)
  m_matrixcont<-matrix(as.vector(m$country),nrow=males,ncol=len,byrow=T)
  m_framcont<-data.frame(m_matrixcont)
  m_contcol<-c(m_framcont$X1)

  mHGDP.pca <- prcomp(m_fram, center = TRUE, scale = TRUE)
  
  
  library(ggbiplot)
  f_PCA <- ggbiplot(fHGDP.pca, var.axes=FALSE, groups = f_contcol, ellipse = TRUE)+ ggtitle("Females")+ theme(plot.title = element_text(size = 8, hjust = 0.5)) 
  m_PCA <- ggbiplot(mHGDP.pca, var.axes=FALSE, groups = m_contcol, ellipse = TRUE)+ ggtitle("Males")+ theme(plot.title = element_text(size = 8, hjust = 0.5)) 
  
figure <- ggarrange(f_PCA, m_PCA, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))

annotate_figure(figure, top = text_grob(title, color = "black", size = 20), fig.lab = "")
}
```

# PCA for the most variable TEs

My first idea is to compare the PCA for the `copynumber` of all the **repetitive sequences** (RepSeq) in the dataset with a more specific PCA, the one considering only the most **variable** RepSeq. To create this subset I used the same **filters** applied in scripts 1 and 2.
```{r}
TE_cutoff <- filter(HGDPcutoff, type=='te') %>% group_by(familyname, sex) %>% mutate(max=max(copynumber), min=min(copynumber)) %>% mutate(diff = max-min, ratio = max/min)

out_abs <- filter(TE_cutoff, diff>200 & diff<Inf)
out_rel <- filter(TE_cutoff, ratio>5 & ratio<Inf & max>1.5)
out_abs_names <- c(unique(out_abs$familyname))
out_rel_names <- c(unique(out_rel$familyname))
out_names <- c(out_abs_names, out_rel_names[!(out_rel_names %in% out_abs_names)]) %>% sort()

most_variable<-subset(HGDPcutoff, type=="te") %>% filter(familyname %in% out_names)

PCA(HGDPcutoff, "All the repetitive sequences")
```

* **Africans** are clearly separated from the others.
* **Eurasians**, which comprehend `Europe`, `Middle_East`, `Central_South_Asia` and `East_Asia`, are always clustered together. This makes sense because there is no strong geographic barrier among these populations, thus there was a strong gene flow across thousands of years which led to a more homogeneous genetic background.
* **Oceanians** are also a separated group, with some individuals closer to Africans and others to Eurasians.
* **Americans** are clustered together, slightly separated from Eurasians.

All the described patterns are more evident in **females**. In the males, there are some individuals from America and Eurasia that are clustered together with the africans. This was previously noted in Scripts 1-2.

```{r}
PCA(most_variable, "Most variable repetitive sequences")
```

Filtering for the most variable sequences does not change the figure so much. I only notice that the percentage of **explained variability** has increased for both the plots and both axes.


# PCA for different RepSeq families

```{r}
classification_tot <- read_tsv("/Users/rpianezza/TE/ric-documentation-Rmd/other-files/repbase_classification.txt", col_names = c("familyname", "superfamily", "shared_with"))

families_in_HGDP <- filter(HGDPcutoff, type=="te") %>% distinct(familyname) %>% pull()

classification <- filter(classification_tot, familyname %in% families_in_HGDP)

HGDP_class <- full_join(HGDPcutoff, classification, by="familyname")

write_tsv(HGDP_class, "/Volumes/Temp1/rpianezza/TE/summary-HGDP/HGDP_cutoff_classified.tsv", col_names = TRUE)

DNA_names <- c("Crypton", "hAT", "Helitron", "Kolobok", "Mariner/Tc1", "Merlin", "MuDR", "piggyBac", "DNA transposon")
LINE_names <- c("L1", "CR1", "L2", "Crack", "RTE", "RTEX", "R4", "Vingi", "Tx1", "Penelope")
SINE_names <- c("SINE1/7SL", "SINE2/tRNA", "SINE3/5S", "SINE")
LTR_names <- c("ERV1", "ERV2", "ERV3", "Gypsy", "Endogenous Retrovirus", "LTR Retrotransposon", "Long terminal repeat", "Non-LTR Retrotransposon")
satellites_names <- c("Satellite", "satellite", "SAT")

(classification <- HGDP_class %>% mutate(class = case_when(superfamily %in% DNA_names ~ "DNA", superfamily %in% LINE_names ~ "LINE", superfamily %in% SINE_names ~ "SINE", superfamily %in% LTR_names ~ "LTR", superfamily %in% satellites_names ~ "satellite")))
```
This part of the code is work in progress. My idea was to give a bit of context to each RepSeq in the dataset, adding a column with the family of the sequence. Anyway, I did not find a way to do that apart from manually annotate the family of each TE, using the info available in RepBase. I know that this approach is not only super slow but also error prone, so I will work on that to improve its reliability.

```{r}
LINE <- filter(classification, type=="te", class=="LINE")
PCA(LINE, "LINEs")

DNA <- filter(classification, type=="te", class=="DNA")
PCA(DNA, "DNA transposons")

LTR <- filter(classification, type=="te", class=="LTR")
PCA(LTR, "LTR retrotransposons")

simple_repeats <- filter(classification, type=="te", class=="satellite")
PCA(simple_repeats, "Satellites")
```

I notice that the pattern previously described is evident in **non-LTR retrotransposons** as well as in **DNA transposons**, but not in **simple repeats**. We do not expect simple repeats to rapidly spread into different populations as TEs, so this is an expected result, but still nice to see.

```{r}
L1 <- filter(classification, type=="te", superfamily=="L1")
PCA(L1, "LINE-1 retrotransposons")

L2 <- subset(classification, type=="te") %>% filter(grepl("L2|L3|L4",familyname))
PCA(L2, "LINE-2/3/4 retrotransposons")
```

The same pattern is well shown in **LINE-1** retrotransposons, known to be the most active TE family in humans, but is totally absent in **LINE-2/3-4** retrotransposons, which are expected to be extint in humans.

