---
title: "Analyzing the TEs with the highest relative or absolute difference between their minimum and maximum abundance estimate"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/human-data')
```

# Unbiased Population Variation of Human Transposable Elements - Script 7
This is the seventh out of eight scripts describing the creation and analysis of the dataset of human TE abundance.
This script works with the version of the HGDP dataset created in Script 2.
Additionally, we perform all analyses seperately for males and females, as we established there are differences between the sexes in Script 5.
```{r}
library(tidyverse)
#setwd(~/human-data)
HGDPcutoff<-read_delim("summary-HGDP/USEME_HGDP_complete_reflib6.2_mq10_batchinfo_cutoff0.01.txt",comment="#")
names(HGDPcutoff)<-c("ID","Pop","sex","Country","type","familyname","length","reads","copynumber","batch")

ftecutoff<-subset(HGDPcutoff, sex=="female" & type=="te")
mtecutoff<-subset(HGDPcutoff, sex=="male" & type=="te")
```

The idea of this code is to plot the general distribution of TE abundance estimates in a boxplot.
This means that we will display the 25% and 75% quantile and the median in a box, and show the outlier dots above and below this median.
Looking at this plot for all 965 TEs seems excessive, so I did want to create a subset. However, due to the data structure, this was not as trivial as before. I found a workaround, but I do believe it is not very intuitive. I will do my best to explain my code and the idea behind it.
If you can think of a cleaner way to do this, feel free to implement it.
As the analysis is done for males and females separately, I will try to show the code in detail for the female analysis, and then add the full code for the male analysis in the end.
First, I wanted to identify the minimum, mean and maximum value for each TE sequence.
This code chunk creates a data frame containing respectively the name of the TE and the minimum, mean and maximum value for each respective TE.
```{r}
tenames <- unique(ftecutoff$familyname)
femaleminmeanmax<-c()
for (i in tenames){
  femaleminmeanmax<-c(femaleminmeanmax,min(ftecutoff$copynumber[ftecutoff$familyname==i]),mean(ftecutoff$copynumber[ftecutoff$familyname==i]),max(ftecutoff$copynumber[ftecutoff$familyname==i]))}
femaleminmeanmaxmatrix<-matrix(femaleminmeanmax,ncol=3,byrow=T)
femaleminmeanmaxframesepcols<-data.frame(tenames,femaleminmeanmaxmatrix)
names(femaleminmeanmaxframesepcols)<-c('tenames','min','mean','max')
```
This analysis, similar to the comparison of mean abundance differences, was done for both a relative and an absolute comparison.
I personally think that the relative comparison is likely the more informative, but I include the code for both.

### Relative comparison

#### Females
From the whole dataset, we want to create a subset containing only the TEs with the highest relative differences between minimum and maximum value.
So what I did was to create a subset containing only the names of TEs with a specific fold difference between the minimum and the maximum copy number value.
In this specific example, i chose a fourfold difference. If you want to include more/less TEs in the plot, feel free to change the fold value.
I also included the condition that the value should be lower than Infinite as otherwise this caused some issues.
```{r}
outliertenames<-femaleminmeanmaxframesepcols$tenames[femaleminmeanmaxframesepcols$max/femaleminmeanmaxframesepcols$min>4 & femaleminmeanmaxframesepcols$max/femaleminmeanmaxframesepcols$min<Inf]
```
For the production of the plot we do want all values of the respective TE, and not just its minimum and maxmimum value. Thus, we need to create a subset of the full dataset containing only the TEs with the highest fold difference between minimum and maximum value. As we already have a vector containing the names of the TEs we want for the female dataset, we just need to subset the dataset. Then we order the dataset and make sure that the order is correctly displayed in the follwoing plot.
```{r}
fteoutlier<-ftecutoff[ftecutoff$familyname %in% outliertenames,]
fteoutlierorder<-fteoutlier[order(fteoutlier$copynumber,decreasing=T),]
fteoutlierorder$familyname<-factor(fteoutlierorder$familyname,levels=unique(fteoutlierorder$familyname))
```
Now we have a dataset containing only the TEs with the highest differences in abundance between minimum and maximum value.
We can now create a simple boxplot to look at the distribution of these values.
```{r}
ggplot(fteoutlierorder,aes(x=familyname,y=log(copynumber)))+geom_boxplot(notch=F)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) #Most of this is pretty uninteresting. How do I subset for the most variable TEs instead of the most abundant?
```
One quite exciting thing about this plot is the L1 variant L1ME5.
This TE was already behaving quite interestingly in the male-female comparison and it does again not disappoint here.
It gets especially interesting if you look at where the values with markedly lower abundance seen in the plot actually come from
```{r}
L1ME5<-fteoutlierorder[fteoutlierorder$familyname=='L1ME5',]
L1ME5[L1ME5$copynumber<5,]
```
As you can see in this table, nearly all of the low abundance values come from African populations.
Which I think is quite cool a finding, and I presume future analyses will look at this abundance difference in more detail.



#### Males
The analysis for males is understandably mirroring the one for females, so I will not explain the code in too much detail.
I used a fivefold cutoff in this case to have a roughly similar number of TEs as in the plot for females, but again, these cutoffs should probably be changed and played around with a bit.
```{r}
maleminmeanmax<-c()
for (i in tenames){
  maleminmeanmax<-c(maleminmeanmax,min(mtecutoff$copynumber[mtecutoff$familyname==i]),mean(mtecutoff$copynumber[mtecutoff$familyname==i]),max(mtecutoff$copynumber[mtecutoff$familyname==i]))}
maleminmeanmaxmatrix<-matrix(maleminmeanmax,ncol=3,byrow=T)
maleminmeanmaxframesepcols<-data.frame(tenames,maleminmeanmaxmatrix)
names(maleminmeanmaxframesepcols)<-c('tenames','min','mean','max')

outliertenames<-maleminmeanmaxframesepcols$tenames[maleminmeanmaxframesepcols$max/maleminmeanmaxframesepcols$min>5 & maleminmeanmaxframesepcols$max/maleminmeanmaxframesepcols$min<Inf]
mteoutlier<-mtecutoff[mtecutoff$familyname %in% outliertenames,]
mteoutlierorder<-mteoutlier[order(mteoutlier$copynumber,decreasing=T),]
mteoutlierorder$familyname<-factor(mteoutlierorder$familyname,levels=unique(mteoutlierorder$familyname))

ggplot(mteoutlierorder,aes(x=familyname,y=log(copynumber)))+geom_boxplot(notch=F)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```
Notabvly, the results specific to L1ME5 seem not nearly as clear, and low copy number estimates exist in many other Continents.
Definitely worth more investigation!
```{r}
L1ME5<-mteoutlierorder[mteoutlierorder$familyname=='L1ME5',]
L1ME5[L1ME5$copynumber<3,]
```

### Absolute comparison

#### Females
We use the same dataset for the mean, minimum and maximum values as those do not change.
The only thing that changes is our definition of outliers. Instead of a relative fold-difference between the minimum and maximum value, we now want to use an absolute value of maximum copy number vs minimum copy number as a threshold.
Again, experimenting with the value will lead to the inclusion of more or less TEs in the plot.
```{r}
outliertenames<-femaleminmeanmaxframesepcols$tenames[femaleminmeanmaxframesepcols$max-femaleminmeanmaxframesepcols$min>200 & femaleminmeanmaxframesepcols$max-femaleminmeanmaxframesepcols$min<Inf]
fteoutlier<-ftecutoff[ftecutoff$familyname %in% outliertenames,]
fteoutlierorder<-fteoutlier[order(fteoutlier$copynumber,decreasing=T),]
fteoutlierorder$familyname<-factor(fteoutlierorder$familyname,levels=unique(fteoutlierorder$familyname))
ggplot(fteoutlierorder,aes(x=familyname,y=log(copynumber)))+geom_boxplot(notch=F)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```


#### Males


```{r}
outliertenames<-maleminmeanmaxframesepcols$tenames[maleminmeanmaxframesepcols$max-maleminmeanmaxframesepcols$min>200 & maleminmeanmaxframesepcols$max-maleminmeanmaxframesepcols$min<Inf]
mteoutlier<-mtecutoff[mtecutoff$familyname %in% outliertenames,]
mteoutlierorder<-mteoutlier[order(mteoutlier$copynumber,decreasing=T),]
mteoutlierorder$familyname<-factor(mteoutlierorder$familyname,levels=unique(mteoutlierorder$familyname))
ggplot(mteoutlierorder,aes(x=familyname,y=log(copynumber)))+geom_boxplot(notch=F)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```



My general issue with this whole series of plots is that we show the TEs with the highest differences, but do these actually represent the TEs that are most variable within/between the populations? And if not, what is a better analyses to focus on the TEs whose variations in copy numbers are the most interesting?