---
title: "SNPs - comparative analysis"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

In this document I would like to summarize the analysis that I have done on repetitive sequences SNPs in the HGDP and in the SGDP. The goal is to better organize and compare the analysis, while using a unique R code for all the plots to avoid errors.

## Prepare the environment

```{r env}
library(tidyverse)
library(ggpubr)
```

## Read the summary files

Here we create 6 R objects containing the summary files (sample names and metadata) for:

* **HGDP**: all the HGDP samples
* **SGDP**: all the SGDP samples
* **HGDP_pcr_free**: the pcr-free HGDP samples
* **SGDP_pcr_free**: the pcr-free SGDP samples
* **HGDP_pcr**: the pcr HGDP samples
* **SGDP_pcr**: the pcr SGDP samples

```{r summary-files}
HGDP<-read_delim("/Volumes/Temp1/rpianezza/TE/summary-HGDP/USEME_HGDP_complete_reflib6.2_mq10_batchinfo_cutoff0.01.txt")
names(HGDP)<-c("ID","pop","sex","country","type","familyname","length","reads","copynumber","batch")

order <- c("Africa","America","Central Asia and Siberia","East Asia","West Eurasia","South Asia","Oceania")
SGDP <- read_tsv("/Volumes/Temp2/rpianezza/SGDP/summary/USEME_SGDP_cutoff") %>% dplyr::rename(ID=biosample)

HGDP_pcr_free_samples <- read_tsv("/Volumes/Temp1/rpianezza/investigation/HGDP-no-PCR/HGDP-only-pcr-free-samples.tsv", col_names = "ID")
HGDP_pcr_free <- HGDP %>% filter(ID %in% HGDP_pcr_free_samples$ID)
HGDP_pcr <- HGDP %>% filter(!(ID %in% HGDP_pcr_free_samples$ID))

SGDP_pcr_free_samples <- read_tsv("/Volumes/Temp1/rpianezza/SGDP/ric-documentation/SGDP-no-PCR.tsv")
SGDP_pcr_free <- SGDP %>% filter(ID %in% SGDP_pcr_free_samples$ID)
SGDP_pcr <- SGDP %>% filter(!(ID %in% SGDP_pcr_free_samples$ID))
```

## Read the SNPs matrixes

The SNPs matrixes are created using the script **frequency_matrix_v2.py** with arguments **--max_freq 0.8 --min_cov 5000 --transv n**. The last argument make the script only considering SNPs which do not create transversions, to avoid GC-bias (only A-T or G-C SNPs are retained).

```{r SNP-matrixes}
HGDP_noGCbias = "/Volumes/Temp1/rpianezza/SNPs-all-analysis/matrixes/HGDP.notransv.08.5000.matrix_processed"
SGDP_noGCbias = "/Volumes/Temp1/rpianezza/SNPs-all-analysis/matrixes/SGDP.notransv.08.5000.matrix_processed"

HGDP_45_55 = "/Volumes/Temp1/rpianezza/SNPs-all-analysis/matrixes/HGDP.08.5000.45-55.matrix_processed"
SGDP_45_55 = "/Volumes/Temp1/rpianezza/SNPs-all-analysis/matrixes/SGDP.08.5000.45-55.matrix_processed"

HGDP_noGCbias_45_55 = "/Volumes/Temp1/rpianezza/SNPs-all-analysis/matrixes/HGDP.notransv.08.5000.45-55.matrix_processed"
SGDP_noGCbias_45_55 = "/Volumes/Temp1/rpianezza/SNPs-all-analysis/matrixes/SGDP.notransv.08.5000.45-55.matrix_processed"
```

## Create the function for PCA plotting

```{r main-function}
PCA <- function(freq_matrix, metadata){

matrix <- read_csv(freq_matrix)
  
f_metadata <- metadata %>% filter(sex=="female") %>% select(ID, sex, country, pop) %>% distinct()
m_metadata <- metadata %>% filter(sex=="male") %>% select(ID, sex, country, pop) %>% distinct()
males_matrix <- inner_join(m_metadata, matrix, by="ID")
females_matrix <- inner_join(f_metadata, matrix, by="ID")

f_pca_data <- females_matrix %>%  select_if(~ !all(. == .[1]))
m_pca_data <- males_matrix %>%  select_if(~ !all(. == .[1]))
f_pca_result <- prcomp(f_pca_data[, -c(1:3)], center = TRUE, scale = TRUE)
m_pca_result <- prcomp(m_pca_data[, -c(1:3)], center = TRUE, scale = TRUE)
      
f_var_explained <- f_pca_result$sdev^2/sum(f_pca_result$sdev^2)
m_var_explained <- m_pca_result$sdev^2/sum(m_pca_result$sdev^2)
      
f <- ggplot(data.frame(f_pca_result$x, country=females_matrix$country), aes(x=PC1,y=PC2, color=country)) + geom_point(size=2, show.legend = FALSE) + labs(x=paste0("PC1: ",round(f_var_explained[1]*100,1),"%"), y=paste0("PC2: ",round(f_var_explained[2]*100,1),"%"))
   
m <- ggplot(data.frame(m_pca_result$x, country=males_matrix$country), aes(x=PC1,y=PC2, color=country)) + geom_point(size=2, show.legend = FALSE) + labs(x=paste0("PC1: ",round(m_var_explained[1]*100,1),"%"), y=paste0("PC2: ",round(m_var_explained[2]*100,1),"%"))
     
plot <- ggarrange(f, m, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))
}

PCA_SGDP <- function(freq_matrix, metadata, order){
  matrix <- read_csv(freq_matrix)
  
  f_metadata <- metadata %>% filter(sex=="female") %>% select(ID, sex, country, pop) %>% distinct()
  m_metadata <- metadata %>% filter(sex=="male") %>% select(ID, sex, country, pop) %>% distinct()
  f_metadata$country_reordered <- factor(f_metadata$country, levels=order)
  m_metadata$country_reordered <- factor(m_metadata$country, levels=order)
  males_matrix <- inner_join(m_metadata, matrix, by="ID")
  females_matrix <- inner_join(f_metadata, matrix, by="ID")

   f_pca_data <- females_matrix %>%  select_if(~ !all(. == .[1]))
   m_pca_data <- males_matrix %>%  select_if(~ !all(. == .[1]))
    f_pca_result <- prcomp(f_pca_data[, -c(1:4)], center = TRUE, scale = TRUE)
    m_pca_result <- prcomp(m_pca_data[, -c(1:4)], center = TRUE, scale = TRUE)
     
    f_var_explained <- f_pca_result$sdev^2/sum(f_pca_result$sdev^2)
    m_var_explained <- m_pca_result$sdev^2/sum(m_pca_result$sdev^2)
        
 f <- ggplot(data.frame(f_pca_result$x, country=females_matrix$country), aes(x=PC1,y=PC2, color=females_matrix$country_reordered)) + geom_point(size=2, show.legend = FALSE) + labs(x=paste0("PC1: ",round(f_var_explained[1]*100,1),"%"), y=paste0("PC2: ",round(f_var_explained[2]*100,1),"%"))
 m <- ggplot(data.frame(m_pca_result$x, country=males_matrix$country), aes(x=PC1,y=PC2, color=males_matrix$country_reordered)) + geom_point(size=2, show.legend = FALSE) + labs(x=paste0("PC1: ",round(m_var_explained[1]*100,1),"%"), y=paste0("PC2: ",round(m_var_explained[2]*100,1),"%"))
       
    plot <- ggarrange(f, m, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))
}

```

```{r HGDP}
# All samples
(HGDP_nobiasSNPs <- PCA(HGDP_noGCbias, HGDP))                                   # no GC biased SNPs
(HGDP_45to55 <- PCA(HGDP_45_55, HGDP))                                          # no GC biased TEs
(HGDP_nobiasSNPs_45to55 <- PCA(HGDP_noGCbias_45_55, HGDP))                      # no GC biased SNPs and TEs

# PCR-free
(HGDP_nobiasSNPs_pcrfree <- PCA(HGDP_noGCbias, HGDP_pcr_free))                  # no GC biased SNPs
(HGDP_45to55_pcrfree <- PCA(HGDP_45_55, HGDP_pcr_free))                         # no GC biased TEs
(HGDP_nobiasSNPs_45to55_pcrfree <- PCA(HGDP_noGCbias_45_55, HGDP_pcr_free))     # no GC biased SNPs and TEs

# PCR
(HGDP_nobias_pcr <- PCA(HGDP_noGCbias, HGDP_pcr))                               # no GC biased SNPs
(HGDP_45to55_pcr <- PCA(HGDP_45_55, HGDP_pcr))                                  # no GC biased TEs
(HGDP_nobiasSNPs_45to55_pcr <- PCA(HGDP_noGCbias_45_55, HGDP_pcr))              # no GC biased SNPs and TEs
```


```{r SGDP}
# All samples
(SGDP_nobiasSNPs <- PCA_SGDP(SGDP_noGCbias, SGDP, order))                                   # no GC biased SNPs
(SGDP_45to55 <- PCA_SGDP(SGDP_45_55, SGDP, order))                                          # no GC biased TEs
(SGDP_nobiasSNPs_45to55 <- PCA_SGDP(SGDP_noGCbias_45_55, SGDP, order))                      # no GC biased SNPs and TEs

# PCR-free
(SGDP_nobiasSNPs_pcrfree <- PCA_SGDP(SGDP_noGCbias, SGDP_pcr_free, order))                  # no GC biased SNPs
(SGDP_45to55_pcrfree <- PCA_SGDP(SGDP_45_55, SGDP_pcr_free, order))                         # no GC biased TEs
(SGDP_nobiasSNPs_45to55_pcrfree <- PCA_SGDP(SGDP_noGCbias_45_55, SGDP_pcr_free, order))     # no GC biased SNPs and TEs
```


