---
title: "HGDP - Geographic distribution by population for the most interesting TEs"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE/')
```

```{r}
library(tidyverse)
library(ggpubr)
library(forcats)

no_pcr_samples <- read_tsv("/Volumes/Temp1/rpianezza/investigation/HGDP-no-PCR/HGDP-only-pcr-free-samples.tsv", col_names = ("ID"))

HGDP <- read_tsv("/Volumes/Temp1/rpianezza/TE/summary-HGDP/USEME_HGDP_mq0_cutoff0.01.txt", col_names = c("ID","pop","sex","country","type","familyname","length","reads","copynumber","batch"), skip=1) %>% mutate(country = recode(country, "Oceania_(SGDP),Oceania"="Oceania")) %>% type_convert() %>% filter(ID %in% no_pcr_samples$ID)
```

# Geographical distribution of interesting TE by population

## Preparatory work

```{r}
coordinates <- read_tsv("/Users/rpianezza/TE/summary-HGDP/HGDP_populationcoordinates.txt", col_names = c("pop", "region", "latitude", "longitude"))
coord <- select(coordinates, pop, latitude, longitude)

TE <- filter(HGDP, type=='te')
```

```{r}
by_pop <- group_by(TE, pop, country, familyname, sex) %>% dplyr::summarise(sd=sd(copynumber), copynumber = mean(copynumber), count=n())
data <- inner_join(x = coordinates, y = by_pop, by = "pop")
```

We are now ready to analyze the geographic distribution of the most interesting TE.
I first create a function which I will use to plot each TE family.

```{r}
plot_map <- function(data, famname){
TE <- filter(data, familyname == famname)
world_map = map_data("world")

ggplot() +
  geom_map(
    data = world_map, map = world_map,
    aes(long, lat, map_id = region),
    color = "white", fill = "lightgray", size = 0) +
  geom_point(
    data = TE, aes(longitude, latitude, color = copynumber, size = count)
  ) + geom_errorbar() + theme(legend.position="top") + scale_colour_gradient(low = "green", high = "red") + theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~sex) + ggtitle(famname)}
```

## Details for every interesting TE

### LINEs

```{r}
plot_map(data, "L1PB1")
plot_map(data, "L1PA3")
plot_map(data, "L1PA7_5")
```

```{r}
plot_map(data, "L1PA4")
plot_map(data, "L1")
plot_map(data, "L1PREC1")
plot_map(data, "L1PA16")
plot_map(data, "L1PA6")
plot_map(data, "L1HS")
plot_map(data, "L1PA7")
#plot_map(data, "L1PB2c")
#plot_map(data, "L1PA10")
##plot_map(data, "L1PA8")
#plot_map(data, "L1PREC2")
#plot_map(data, "L1PA15")
#plot_map(data, "L1P_MA2")
#plot_map(data, "L1MC1")
#plot_map(data, "L1PB2")
#plot_map(data, "L1PB4")
```


### SINEs

```{r}
plot_map(data, "ALU")
plot_map(data, "SVA_A")
```

### DNA transposons

```{r}
plot_map(data, "MER2")
```

### Endogenous retroviruses

```{r}
plot_map(data, "MLT2A1")
```

### Satellites

```{r}
plot_map(data, "GSATII")
```


## Copynumber standard deviation by population

```{r}
plot_cn_sd <- function(data, TE, ylimits){
  
m <- filter(data, sex=="male")
f <- filter(data, sex=="female")

m_singleTE <- filter(m, familyname==TE) %>% mutate(pop = fct_reorder(pop, copynumber))
m_plot <- ggplot(m_singleTE, aes(pop, copynumber, fill=country))+
  geom_bar(stat = "identity")+
  geom_errorbar(aes(ymin=copynumber-sd, ymax=copynumber+sd), alpha=0.5)+
  ggtitle("male") + ylab("Mean copynumber") + xlab("Population") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=4), axis.title.y = element_blank()) + coord_cartesian(ylim=ylimits) +
  theme(plot.title = element_text(hjust = 0.5))

f_singleTE <- filter(f, familyname==TE) %>% mutate(pop = fct_reorder(pop, copynumber))
f_plot <- ggplot(f_singleTE, aes(pop, copynumber, fill=country))+
  geom_bar(stat = "identity")+
  geom_errorbar(aes(ymin=copynumber-sd, ymax=copynumber+sd), alpha=0.5)+
  ggtitle("female") + ylab("Mean copynumber") + xlab("Population") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=4)) + coord_cartesian(ylim=ylimits) +
  theme(plot.title = element_text(hjust = 0.5))

figure <- ggarrange(f_plot, m_plot, common.legend = TRUE)
(figure_final <- annotate_figure(figure, top=text_grob(TE)))
}

plot_cn_sd(data, "L1PA16", c(2400, 3200))
plot_cn_sd(data, "L1PA3", c(4000, 5500))
plot_cn_sd(data, "L1PB1", c(1600, 2300))
plot_cn_sd(data, "L1PA7_5", c(1800, 2300))
plot_cn_sd(data, "ALU", c(150000, 220000))
plot_cn_sd(data, "SVA_A", c(2000, 2500))
plot_cn_sd(data, "MER2", c(750, 1050))
plot_cn_sd(data, "MLT2A1", c(1100, 1525))
plot_cn_sd(data, "L2", c(5, 11))
```

