---
title: "HGDP - General analysis about TEs geographic distribution"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE/')
```

In all the previous analysis, we just focused on the most variable TE, selected in the scripts 1-2-3. Here I want to give a more comprehensive overview of the geographic distributions of TE abundance.

## Creating a customized global classification of TE

First, I read the file */Volumes/Temp1/rpianezza/TE/summary-HGDP/HGDP_cutoff_classified.tsv*, which is a file which combine the previously used *HGDP_cutoff* with *repbase_classification.txt*, a file containing the classification of each TE present in our reference library, directly from RepBase.
```{r}
library(tidyverse)
library(ggpubr)

no_pcr_samples <- read_tsv("/Volumes/Temp1/rpianezza/investigation/HGDP-no-PCR/HGDP-only-pcr-free-samples.tsv", col_names = ("ID"))

HGDP <- read_tsv("/Volumes/Temp1/rpianezza/TE/summary-HGDP/HGDP_cutoff_classified.tsv", col_names = c("ID","pop","sex","country","type","familyname","length","reads","copynumber","batch", "superfamily", "shared_with"), skip=1) %>% type_convert() %>% filter(ID %in% no_pcr_samples$ID)
```

Then I add the coordinates for each population and create the dataframe `data` which can be plotted using the funcion `plot_map` created in script 4. This function can be used to check the results of the following analysis in details.
```{r}
coordinates <- read_tsv("/Users/rpianezza/TE/summary-HGDP/HGDP_populationcoordinates.txt", col_names = c("pop", "region", "latitude", "longitude"))
coord <- select(coordinates, pop, latitude, longitude)

by_pop <- group_by(HGDP, pop, country, familyname, type, sex) %>% dplyr::summarise(sd=sd(copynumber), copynumber = mean(copynumber), count=n())
data <- inner_join(coord, by_pop, by = "pop")
```

The classification available in RepBase give us the `superfamily` of each `familyname`. This helps us to create a further larger classification which divides the repetitive sequences in the library into 5 main classes:

* DNA transposons (`DNA`)
* LINE (`LINE`)
* SINE (`SINE`, comprehends SVA_A)
* LTR transposons (`LTR`, all the endogenous retroviruses)
* Satellites (`satellite`)

Note that some families are not included in this classification. As reported in <https://mobilednajournal.biomedcentral.com/articles/10.1186/s13100-017-0107-y#Tab2>, some TE are still `non classified` (`NA`).
```{r}
DNA_names <- c("Crypton", "hAT", "Helitron", "Kolobok", "Mariner/Tc1", "Merlin", "MuDR", "piggyBac", "DNA transposon")
LINE_names <- c("L1", "CR1", "L2", "Crack", "RTE", "RTEX", "R4", "Vingi", "Tx1", "Penelope")
SINE_names <- c("SINE1/7SL", "SINE2/tRNA", "SINE3/5S", "SINE")
LTR_names <- c("ERV1", "ERV2", "ERV3", "Gypsy", "Endogenous Retrovirus", "LTR Retrotransposon", "Long terminal repeat", "Non-LTR Retrotransposon")
satellites_names <- c("Satellite", "satellite", "SAT")

classification <- select(HGDP, familyname, superfamily) %>% mutate(class = case_when(superfamily %in% DNA_names ~ "DNA", superfamily %in% LINE_names ~ "LINE", superfamily %in% SINE_names ~ "SINE", superfamily %in% LTR_names ~ "LTR", superfamily %in% satellites_names ~ "satellite"))
```

## Geographic classification and mean abundance by continent

After creating the sequence classification, I create also a simplified **geographic classification**. All the populations from `Central South Asia`, `East Asia`, `Middle East`, and `Europe` are grouped under the label `Eurasia`, since from the PCA and from all the other analysis these populations always clustered together in terms of TE abundance.

Then, I calculate the **mean copynumber** of each TE for every continent (`country`).
```{r}
(continents <- by_pop %>% group_by(country, familyname, type, sex) %>% mutate(country = replace(country, country == 'Central_South_Asia', 'Eurasia')) %>% mutate(country = replace(country, country == 'East_Asia', 'Eurasia')) %>% mutate(country = replace(country, country == 'Middle_East', 'Eurasia')) %>% mutate(country = replace(country, country == 'Europe', 'Eurasia')) %>% dplyr::summarise(country_mean=mean(copynumber)) %>% arrange(familyname) %>% pivot_wider(names_from = country, values_from = country_mean) %>% full_join(classification, by="familyname") %>% distinct())
```

## Functions

This code chunk contains a function `pie_chart_more_in` which creates two pie charts that shows the percentage of TE families that are more abundant in each continent, one for the females and one for the males.

The two arguments are:

* `class_or_superfamily`: used to subset the dataset and select only a class or a superfamily to show in the pie charts. If "class", it will subset looking for TE which are in a class, if "superfamily" it will do the same but just in one superfamily
* `which_TE`: the specific `class` or `superfamily` name is specified in this argument (es. "LINE").

```{r}
pie_chart_more_in <- function(data, class_or_superfamily, which_TE){

if (class_or_superfamily=="class"){subset <- filter(data, class==which_TE)}
if (class_or_superfamily=="superfamily"){subset <- filter(data, superfamily==which_TE)}
if (class_or_superfamily=="type"){subset <- filter(data, type==which_TE)}
  
f_eurasia <- filter(subset, Eurasia-Africa>0, Eurasia-America>0, Eurasia-Oceania>0, sex=="female") %>% ungroup %>% dplyr::summarise(eurasia = n()) %>% pull
f_africa <- filter(subset, Africa-Eurasia>0, Africa-America>0, Africa-Oceania>0, sex=="female") %>% ungroup %>% dplyr::summarise(africa = n()) %>% pull
f_america <- filter(subset, America-Eurasia>0, America-Africa>0, America-Oceania>0, sex=="female") %>% ungroup %>% dplyr::summarise(america = n()) %>% pull
f_oceania <- filter(subset, Oceania-Eurasia>0, Oceania-Africa>0, Oceania-America>0, sex=="female") %>% ungroup %>% dplyr::summarise(oceania = n()) %>% pull

continents_names <- c("Eurasia", "Africa", "America", "Oceania")
f_morein <- c(f_eurasia, f_africa, f_america, f_oceania)

tot=f_eurasia+f_africa+f_america+f_oceania

f_morein_tibble <- tibble(continents_names, f_morein) %>% mutate(f_perc=round((f_morein/tot), 2))

f_bp <- ggplot(f_morein_tibble, aes("More abundant in", y=f_morein, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

f_pie <- f_bp + coord_polar("y", start=0) +
  ggtitle("Females") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = f_morein), position = position_stack(vjust = 0.5), show.legend = FALSE)


m_eurasia <- filter(subset, Eurasia-Africa>0, Eurasia-America>0, Eurasia-Oceania>0, sex=="male") %>% ungroup %>% dplyr::summarise(eurasia = n()) %>% pull
m_africa <- filter(subset, Africa-Eurasia>0, Africa-America>0, Africa-Oceania>0, sex=="male") %>% ungroup %>% dplyr::summarise(africa = n()) %>% pull
m_america <- filter(subset, America-Eurasia>0, America-Africa>0, America-Oceania>0, sex=="male") %>% ungroup %>% dplyr::summarise(america = n()) %>% pull
m_oceania <- filter(subset, Oceania-Eurasia>0, Oceania-Africa>0, Oceania-America>0, sex=="male") %>% ungroup %>% dplyr::summarise(oceania = n()) %>% pull

m_morein <- c(m_eurasia, m_africa, m_america, m_oceania)

m_morein_tibble <- tibble(continents_names, m_morein) %>% mutate(m_perc=round((m_morein/tot), 2))

m_bp <- ggplot(m_morein_tibble, aes("More abundant in", y=m_morein, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

m_pie <- m_bp + coord_polar("y", start=0) +
  ggtitle("Males") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = m_morein), position = position_stack(vjust = 0.5), show.legend = FALSE)

figure <- ggarrange(f_pie, m_pie, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))

text <- paste(which_TE, "- How many families are most abundant in each continent?")

(figure_final <- annotate_figure(figure, top = text_grob(text, color = "black", size = 15, hjust = 0.55, vjust = 2), fig.lab = ""))

}
```

This second function `pie_chart_less_in` does the same thing as the one above, but it will check for the `less abundant` TE in each continent instead.
```{r}
pie_chart_less_in <- function(data, class_or_superfamily, which_TE){

if (class_or_superfamily=="class"){subset <- filter(data, class==which_TE)}
if (class_or_superfamily=="superfamily"){subset <- filter(data, superfamily==which_TE)}
if (class_or_superfamily=="type"){subset <- filter(data, type==which_TE)}

less_f_eurasia <- filter(subset, Eurasia-Africa<0, Eurasia-America<0, Eurasia-Oceania<0, sex=="female") %>% ungroup %>% dplyr::summarise(eurasia = dplyr::n()) %>% pull
less_f_africa <- filter(subset, Africa-Eurasia<0, Africa-America<0, Africa-Oceania<0, sex=="female") %>% ungroup %>% dplyr::summarise(africa = dplyr::n()) %>% pull
less_f_america <- filter(subset, America-Eurasia<0, America-Africa<0, America-Oceania<0, sex=="female") %>% ungroup %>% dplyr::summarise(america = dplyr::n()) %>% pull
less_f_oceania <- filter(subset, Oceania-Eurasia<0, Oceania-Africa<0, Oceania-America<0, sex=="female") %>% ungroup %>% dplyr::summarise(oceania = dplyr::n()) %>% pull

f_lessin <- c(less_f_eurasia, less_f_africa, less_f_america, less_f_oceania)
continents_names <- c("Eurasia", "Africa", "America", "Oceania")
tot=less_f_eurasia+less_f_africa+less_f_america+less_f_oceania

f_lessin_tibble <- tibble(continents_names, f_lessin) %>% mutate(less_f_perc=round((f_lessin/tot), 2))

less_f_bp <- ggplot(f_lessin_tibble, aes("Less abundant in", y=f_lessin, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

less_f_pie <- less_f_bp + coord_polar("y", start=0) +
  ggtitle("Females") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = f_lessin), position = position_stack(vjust = 0.5), show.legend = FALSE)


less_m_eurasia <- filter(subset, Eurasia-Africa<0, Eurasia-America<0, Eurasia-Oceania<0, sex=="male") %>% ungroup %>% dplyr::summarise(eurasia = dplyr::n()) %>% pull
less_m_africa <- filter(subset, Africa-Eurasia<0, Africa-America<0, Africa-Oceania<0, sex=="male") %>% ungroup %>% dplyr::summarise(africa = dplyr::n()) %>% pull
less_m_america <- filter(subset, America-Eurasia<0, America-Africa<0, America-Oceania<0, sex=="male") %>% ungroup %>% dplyr::summarise(america = dplyr::n()) %>% pull
less_m_oceania <- filter(subset, Oceania-Eurasia<0, Oceania-Africa<0, Oceania-America<0, sex=="male") %>% ungroup %>% dplyr::summarise(oceania = dplyr::n()) %>% pull

m_lessin <- c(less_m_eurasia, less_m_africa, less_m_america, less_m_oceania)

m_lessin_tibble <- tibble(continents_names, m_lessin) %>% mutate(less_m_perc=round((m_lessin/tot), 2))

less_m_bp <- ggplot(m_lessin_tibble, aes("Less abundant in", y=m_lessin, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

less_m_pie <- less_m_bp + coord_polar("y", start=0) +
  ggtitle("Males") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = m_lessin), position = position_stack(vjust = 0.5), show.legend = FALSE)

figure <- ggarrange(less_f_pie, less_m_pie, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))

text <- paste(which_TE, "- How many families are least abundant in each continent?")

(figure_final <- annotate_figure(figure, top = text_grob(text, color = "black", size = 15, hjust = 0.55, vjust = 2), fig.lab = ""))
}
```


```{r}
chi_most <- function(data, class_or_superfamily, which_TE){
  
if (class_or_superfamily=="class"){subset <- filter(data, class==which_TE)}
if (class_or_superfamily=="superfamily"){subset <- filter(data, superfamily==which_TE)}
if (class_or_superfamily=="type"){subset <- filter(data, type==which_TE)}

chi_m <- ungroup(subset) %>% mutate(most = case_when((Eurasia-Africa>0 & Eurasia-America>0 & Eurasia-Oceania>0) ~ "Eurasia", (Africa-Eurasia>0 & Africa-America>0 & Africa-Oceania>0) ~ "Africa", (America-Africa>0 & America-Eurasia>0 & America-Oceania>0) ~ "America", (Oceania-Africa>0 & Oceania-America>0 & Oceania-Eurasia>0) ~ "Oceania")) %>% group_by(most) %>% dplyr::summarise(n = dplyr::n()) %>% select(n) %>% pull

while (length(chi_m)!=4) {
  chi_m <- append(chi_m, 0)
  }
  
chisq.test(chi_m)
}
```

```{r}
chi_least <- function(data, class_or_superfamily, which_TE){
  
if (class_or_superfamily=="class"){subset <- filter(data, class==which_TE)}
if (class_or_superfamily=="superfamily"){subset <- filter(data, superfamily==which_TE)}
if (class_or_superfamily=="type"){subset <- filter(data, type==which_TE)}

chi_l <- ungroup(subset) %>% mutate(least = case_when((Eurasia-Africa<0 & Eurasia-America<0 & Eurasia-Oceania<0) ~ "Eurasia", (Africa-Eurasia<0 & Africa-America<0 & Africa-Oceania<0) ~ "Africa", (America-Africa<0 & America-Eurasia<0 & America-Oceania<0) ~ "America", (Oceania-Africa<0 & Oceania-America<0 & Oceania-Eurasia<0) ~ "Oceania")) %>% group_by(least) %>% dplyr::summarise(n = dplyr::n()) %>% select(n) %>% pull

while (length(chi_l)!=4) {
  chi_l <- append(chi_l, 0)
  }
  
chi_l
}
```

After creating the functions, we can explore the dataset using them.

## All

```{r}
pie_chart_more_in(continents, "type", "te")
chi_most(continents, "type", "te")

pie_chart_less_in(continents, "type", "te")
#chi_least(continents, "type", "te")
```

## By class

```{r}
pie_chart_more_in(continents, "class", "LINE")
chi_most(continents, "class", "LINE")

pie_chart_less_in(continents, "class", "LINE")
#chi_least(continents, "class", "LINE")
```

* **More abundant**: we have a clear bias towards **Eurasia**. **Africa** follows, while very few TEs are more abundant in the other two continents.

* **Less abundant**: as expected, mainly in **Oceania** and **America**. **Males** from **Eurasia** also shows some low abundance TEs, which may be interesting.

```{r}
#pie_chart_more_in(continents, "class", "SINE")
#chi_most(continents, "class", "SINE")

#pie_chart_less_in(continents, "class", "SINE")
#chi_least(continents, "class", "SINE")
```
The number of **SINE** families is very low, so I do not expect these charts to be relevant.

* **More abundant**: **Eurasia** and **Africa**. In the **males** also **America**.
* **Less abundant**: **Oceania**, followed by `**Africa**.

```{r}
pie_chart_more_in(continents, "class", "DNA")
chi_most(continents, "class", "DNA")

pie_chart_less_in(continents, "class", "DNA")
#chi_least(continents, "class", "DNA")
```

**DNA transposons** shows surpisingly similar plots to **LINEs**!

```{r}
pie_chart_more_in(continents, "class", "LTR")
chi_most(continents, "class", "LTR")

pie_chart_less_in(continents, "class", "LTR")
#chi_least(continents, "class", "LTR")
```

**LTR transposons** are more abundant in **Africa**! **Eurasia** follows.
Less abundant, as expected, in **America** and a bit in **Oceania**. Again, it is interesting to notice how **eurasian males** are very present in the **less abundant** chart.

```{r}
#pie_chart_more_in(continents, "class", "satellite")
#chi_most(continents, "class", "satellite")

#pie_chart_less_in(continents, "class", "satellite")
#chi_least(continents, "class", "satellite")
```
The number of **satellite** families is very low, so I do not expect these charts to be relevant.

* **More abundant**: Evenly distributed apart from **Eurasia** in **females** and **America** in males, which are poorly represented.
* **Less abundant**: Mostly evenly distributed apart from a positive bias towards **America**  in females.

## By superfamily

```{r}
pie_chart_more_in(continents, "superfamily", "L1")
chi_most(continents, "superfamily", "L1")

pie_chart_less_in(continents, "superfamily", "L1")
#chi_least(continents, "superfamily", "L1")
```

LINE-1 (`L1`) are expected to be the most active TEs in humans, thus we expect these charts to be an extremization of the `LINE` plots, containing both L1s and other LINEs (less active). This is exactly what we find here.


## KRAB-ZNF

```{r}
pie_chart_more_in(continents, "type", "krab")
chi_most(continents, "type", "krab")

pie_chart_less_in(continents, "type", "krab")
#chi_least(continents, "type", "krab")
```

# What do we expect with random values?

Here I substitute the TEs `copynumbers` with random values between 0 and 1.

```{r}
set.seed(42)

random_by_pop <- group_by(HGDP, pop, country, familyname, sex) %>% select(!(copynumber)) %>% add_column(copynumber = runif(nrow(.))) %>% dplyr::summarise(sd=sd(copynumber), copynumber = mean(copynumber), count=n())
random_data <- inner_join(coord, random_by_pop, by = "pop")

random_continents <- random_by_pop %>% group_by(country, familyname, sex) %>% mutate(country = replace(country, country == 'Central_South_Asia', 'Eurasia')) %>% mutate(country = replace(country, country == 'East_Asia', 'Eurasia')) %>% mutate(country = replace(country, country == 'Middle_East', 'Eurasia')) %>% mutate(country = replace(country, country == 'Europe', 'Eurasia')) %>% dplyr::summarise(country_mean=mean(copynumber)) %>% arrange(familyname) %>% pivot_wider(names_from = country, values_from = country_mean) %>% inner_join(classification, by="familyname") %>% distinct()

#pie_chart_more_in(random_continents, "class", "LINE")
#pie_chart_less_in(random_continents, "class", "LINE")
#pie_chart_more_in(random_continents, "class", "SINE")
#pie_chart_less_in(random_continents, "class", "SINE")
#pie_chart_more_in(random_continents, "class", "LTR")
#pie_chart_less_in(random_continents, "class", "LTR")
#pie_chart_more_in(random_continents, "class", "DNA")
#pie_chart_less_in(random_continents, "class", "DNA")
#pie_chart_more_in(random_continents, "class", "satellite")
#pie_chart_less_in(random_continents, "class", "satellite")
```


# Removing non variant TE

```{r}
var_names <- HGDP %>% filter(type=="te") %>% group_by(familyname) %>% filter(max(copynumber)-min(copynumber)>1) %>% select(familyname) %>% distinct() %>% as_vector()

variant <- filter(continents, familyname %in% var_names)

#pie_chart_more_in(variant, "class", "LINE")
#chi_most(variant, "class", "LINE")

#pie_chart_less_in(variant, "class", "LINE")
#chi_least(variant, "class", "LINE")

#pie_chart_more_in(variant, "class", "DNA")
#chi_most(variant, "class", "DNA")

#pie_chart_less_in(variant, "class", "DNA")
#chi_least(variant, "class", "DNA")

#pie_chart_more_in(variant, "class", "LTR")
#chi_most(variant, "class", "LTR")

#pie_chart_less_in(variant, "class", "LTR")
#chi_least(variant, "class", "LTR")
```


# Split eurasians
```{r}
(splitted <- by_pop %>% group_by(country, familyname, type, sex) %>% dplyr::summarise(country_mean=mean(copynumber)) %>% arrange(familyname) %>% pivot_wider(names_from = country, values_from = country_mean) %>% full_join(classification, by="familyname") %>% distinct())
```

```{r}
pie_chart_less_in_splitted <- function(data, class_or_superfamily, which_TE){

if (class_or_superfamily=="class"){subset <- filter(data, class==which_TE)}
if (class_or_superfamily=="superfamily"){subset <- filter(data, superfamily==which_TE)}
if (class_or_superfamily=="type"){subset <- filter(data, type==which_TE)}

less_f_Central_South_Asia <- filter(subset, Central_South_Asia-Africa<0, Central_South_Asia-America<0, Central_South_Asia-Oceania<0, Central_South_Asia-East_Asia<0, Central_South_Asia-Europe<0, Central_South_Asia-Middle_East<0,sex=="female") %>% ungroup %>% dplyr::summarise(Central_South_Asia = dplyr::n()) %>% pull
less_f_East_Asia <- filter(subset, East_Asia-Africa<0, East_Asia-America<0, East_Asia-Oceania<0, East_Asia-Central_South_Asia<0, East_Asia-Europe<0, East_Asia-Middle_East<0,sex=="female") %>% ungroup %>% dplyr::summarise(East_Asia = dplyr::n()) %>% pull
less_f_Europe <- filter(subset, Europe-Africa<0, Europe-America<0, Europe-Oceania<0, Europe-Central_South_Asia<0, Europe-East_Asia<0, Europe-Middle_East<0,sex=="female") %>% ungroup %>% dplyr::summarise(Europe = dplyr::n()) %>% pull
less_f_Middle_East <- filter(subset, Middle_East-Africa<0, Middle_East-America<0, Middle_East-Oceania<0, Middle_East-Central_South_Asia<0, Middle_East-East_Asia<0, Middle_East-Europe<0,sex=="female") %>% ungroup %>% dplyr::summarise(Middle_East = dplyr::n()) %>% pull
less_f_Africa <- filter(subset, Africa-Middle_East<0, Africa-America<0, Africa-Oceania<0, Africa-Central_South_Asia<0, Africa-East_Asia<0, Africa-Europe<0,sex=="female") %>% ungroup %>% dplyr::summarise(Africa = dplyr::n()) %>% pull
less_f_America <- filter(subset, America-Middle_East<0, America-Africa<0, America-Oceania<0, America-Central_South_Asia<0, America-East_Asia<0, America-Europe<0,sex=="female") %>% ungroup %>% dplyr::summarise(America = dplyr::n()) %>% pull
less_f_Oceania <- filter(subset, Oceania-Middle_East<0, Oceania-Africa<0, Oceania-America<0, Oceania-Central_South_Asia<0, Oceania-East_Asia<0, Oceania-Europe<0,sex=="female") %>% ungroup %>% dplyr::summarise(Oceania = dplyr::n()) %>% pull

f_lessin <- c(less_f_Central_South_Asia, less_f_East_Asia, less_f_Europe, less_f_Middle_East, less_f_Africa, less_f_America, less_f_Oceania)
continents_names <- c("Central_South_Asia", "East_Asia", "Europe", "Middle_East", "Africa", "America", "Oceania")
tot=less_f_Central_South_Asia+less_f_East_Asia+less_f_Europe+less_f_Middle_East+less_f_Africa+less_f_America+less_f_Oceania

f_lessin_tibble <- tibble(continents_names, f_lessin) %>% mutate(less_f_perc=round((f_lessin/tot), 2))

less_f_bp <- ggplot(f_lessin_tibble, aes("Less abundant in", y=f_lessin, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

less_f_pie <- less_f_bp + coord_polar("y", start=0) +
  ggtitle("Females") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = f_lessin), position = position_stack(vjust = 0.5), show.legend = FALSE)


less_m_Central_South_Asia <- filter(subset, Central_South_Asia-Africa<0, Central_South_Asia-America<0, Central_South_Asia-Oceania<0, Central_South_Asia-East_Asia<0, Central_South_Asia-Europe<0, Central_South_Asia-Middle_East<0,sex=="male") %>% ungroup %>% dplyr::summarise(Central_South_Asia = dplyr::n()) %>% pull
less_m_East_Asia <- filter(subset, East_Asia-Africa<0, East_Asia-America<0, East_Asia-Oceania<0, East_Asia-Central_South_Asia<0, East_Asia-Europe<0, East_Asia-Middle_East<0,sex=="male") %>% ungroup %>% dplyr::summarise(East_Asia = dplyr::n()) %>% pull
less_m_Europe <- filter(subset, Europe-Africa<0, Europe-America<0, Europe-Oceania<0, Europe-Central_South_Asia<0, Europe-East_Asia<0, Europe-Middle_East<0,sex=="male") %>% ungroup %>% dplyr::summarise(Europe = dplyr::n()) %>% pull
less_m_Middle_East <- filter(subset, Middle_East-Africa<0, Middle_East-America<0, Middle_East-Oceania<0, Middle_East-Central_South_Asia<0, Middle_East-East_Asia<0, Middle_East-Europe<0,sex=="male") %>% ungroup %>% dplyr::summarise(Middle_East = dplyr::n()) %>% pull
less_m_Africa <- filter(subset, Africa-Middle_East<0, Africa-America<0, Africa-Oceania<0, Africa-Central_South_Asia<0, Africa-East_Asia<0, Africa-Europe<0,sex=="male") %>% ungroup %>% dplyr::summarise(Africa = dplyr::n()) %>% pull
less_m_America <- filter(subset, America-Middle_East<0, America-Africa<0, America-Oceania<0, America-Central_South_Asia<0, America-East_Asia<0, America-Europe<0,sex=="male") %>% ungroup %>% dplyr::summarise(America = dplyr::n()) %>% pull
less_m_Oceania <- filter(subset, Oceania-Middle_East<0, Oceania-Africa<0, Oceania-America<0, Oceania-Central_South_Asia<0, Oceania-East_Asia<0, Oceania-Europe<0,sex=="male") %>% ungroup %>% dplyr::summarise(Oceania = dplyr::n()) %>% pull

m_lessin <- c(less_m_Central_South_Asia, less_m_East_Asia, less_m_Europe, less_m_Middle_East, less_m_Africa, less_m_America, less_m_Oceania)
tot=less_m_Central_South_Asia+less_m_East_Asia+less_m_Europe+less_m_Middle_East+less_m_Africa+less_m_America+less_m_Oceania

m_lessin_tibble <- tibble(continents_names, m_lessin) %>% mutate(less_m_perc=round((m_lessin/tot), 2))

less_m_bp <- ggplot(m_lessin_tibble, aes("Less abundant in", y=m_lessin, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

less_m_pie <- less_m_bp + coord_polar("y", start=0) +
  ggtitle("Males") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = m_lessin), position = position_stack(vjust = 0.5), show.legend = FALSE)

figure <- ggarrange(less_f_pie, less_m_pie, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))

text <- paste(which_TE, "- How many families are least abundant in each continent?")

(figure_final <- annotate_figure(figure, top = text_grob(text, color = "black", size = 15, hjust = 0.55, vjust = 2), fig.lab = ""))
}

pie_chart_more_in_splitted <- function(data, class_or_superfamily, which_TE){

if (class_or_superfamily=="class"){subset <- filter(data, class==which_TE)}
if (class_or_superfamily=="superfamily"){subset <- filter(data, superfamily==which_TE)}
if (class_or_superfamily=="type"){subset <- filter(data, type==which_TE)}
  
  
more_f_Central_South_Asia <- filter(subset, Central_South_Asia-Africa>0, Central_South_Asia-America>0, Central_South_Asia-Oceania>0, Central_South_Asia-East_Asia>0, Central_South_Asia-Europe>0, Central_South_Asia-Middle_East>0,sex=="female") %>% ungroup %>% dplyr::summarise(Central_South_Asia = dplyr::n()) %>% pull
more_f_East_Asia <- filter(subset, East_Asia-Africa>0, East_Asia-America>0, East_Asia-Oceania>0, East_Asia-Central_South_Asia>0, East_Asia-Europe>0, East_Asia-Middle_East>0,sex=="female") %>% ungroup %>% dplyr::summarise(East_Asia = dplyr::n()) %>% pull
more_f_Europe <- filter(subset, Europe-Africa>0, Europe-America>0, Europe-Oceania>0, Europe-Central_South_Asia>0, Europe-East_Asia>0, Europe-Middle_East>0,sex=="female") %>% ungroup %>% dplyr::summarise(Europe = dplyr::n()) %>% pull
more_f_Middle_East <- filter(subset, Middle_East-Africa>0, Middle_East-America>0, Middle_East-Oceania>0, Middle_East-Central_South_Asia>0, Middle_East-East_Asia>0, Middle_East-Europe>0,sex=="female") %>% ungroup %>% dplyr::summarise(Middle_East = dplyr::n()) %>% pull
more_f_Africa <- filter(subset, Africa-Middle_East>0, Africa-America>0, Africa-Oceania>0, Africa-Central_South_Asia>0, Africa-East_Asia>0, Africa-Europe>0,sex=="female") %>% ungroup %>% dplyr::summarise(Africa = dplyr::n()) %>% pull
more_f_America <- filter(subset, America-Middle_East>0, America-Africa>0, America-Oceania>0, America-Central_South_Asia>0, America-East_Asia>0, America-Europe>0,sex=="female") %>% ungroup %>% dplyr::summarise(America = dplyr::n()) %>% pull
more_f_Oceania <- filter(subset, Oceania-Middle_East>0, Oceania-Africa>0, Oceania-America>0, Oceania-Central_South_Asia>0, Oceania-East_Asia>0, Oceania-Europe>0,sex=="female") %>% ungroup %>% dplyr::summarise(Oceania = dplyr::n()) %>% pull

f_morein <- c(more_f_Central_South_Asia, more_f_East_Asia, more_f_Europe, more_f_Middle_East, more_f_Africa, more_f_America, more_f_Oceania)
continents_names <- c("Central_South_Asia", "East_Asia", "Europe", "Middle_East", "Africa", "America", "Oceania")
tot=more_f_Central_South_Asia+more_f_East_Asia+more_f_Europe+more_f_Middle_East+more_f_Africa+more_f_America+more_f_Oceania

f_morein_tibble <- tibble(continents_names, f_morein) %>% mutate(more_f_perc=round((f_morein/tot), 2))

more_f_bp <- ggplot(f_morein_tibble, aes("More abundant in", y=f_morein, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

more_f_pie <- more_f_bp + coord_polar("y", start=0) +
  ggtitle("Females") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = f_morein), position = position_stack(vjust = 0.5), show.legend = FALSE)


more_m_Central_South_Asia <- filter(subset, Central_South_Asia-Africa>0, Central_South_Asia-America>0, Central_South_Asia-Oceania>0, Central_South_Asia-East_Asia>0, Central_South_Asia-Europe>0, Central_South_Asia-Middle_East>0,sex=="male") %>% ungroup %>% dplyr::summarise(Central_South_Asia = dplyr::n()) %>% pull
more_m_East_Asia <- filter(subset, East_Asia-Africa>0, East_Asia-America>0, East_Asia-Oceania>0, East_Asia-Central_South_Asia>0, East_Asia-Europe>0, East_Asia-Middle_East>0,sex=="male") %>% ungroup %>% dplyr::summarise(East_Asia = dplyr::n()) %>% pull
more_m_Europe <- filter(subset, Europe-Africa>0, Europe-America>0, Europe-Oceania>0, Europe-Central_South_Asia>0, Europe-East_Asia>0, Europe-Middle_East>0,sex=="male") %>% ungroup %>% dplyr::summarise(Europe = dplyr::n()) %>% pull
more_m_Middle_East <- filter(subset, Middle_East-Africa>0, Middle_East-America>0, Middle_East-Oceania>0, Middle_East-Central_South_Asia>0, Middle_East-East_Asia>0, Middle_East-Europe>0,sex=="male") %>% ungroup %>% dplyr::summarise(Middle_East = dplyr::n()) %>% pull
more_m_Africa <- filter(subset, Africa-Middle_East>0, Africa-America>0, Africa-Oceania>0, Africa-Central_South_Asia>0, Africa-East_Asia>0, Africa-Europe>0,sex=="male") %>% ungroup %>% dplyr::summarise(Africa = dplyr::n()) %>% pull
more_m_America <- filter(subset, America-Middle_East>0, America-Africa>0, America-Oceania>0, America-Central_South_Asia>0, America-East_Asia>0, America-Europe>0,sex=="male") %>% ungroup %>% dplyr::summarise(America = dplyr::n()) %>% pull
more_m_Oceania <- filter(subset, Oceania-Middle_East>0, Oceania-Africa>0, Oceania-America>0, Oceania-Central_South_Asia>0, Oceania-East_Asia>0, Oceania-Europe>0,sex=="male") %>% ungroup %>% dplyr::summarise(Oceania = dplyr::n()) %>% pull

m_morein <- c(more_m_Central_South_Asia, more_m_East_Asia, more_m_Europe, more_m_Middle_East, more_m_Africa, more_m_America, more_m_Oceania)
tot=more_m_Central_South_Asia+more_m_East_Asia+more_m_Europe+more_m_Middle_East+more_m_Africa+more_m_America+more_m_Oceania

m_morein_tibble <- tibble(continents_names, m_morein) %>% mutate(more_m_perc=round((m_morein/tot), 2))

more_m_bp <- ggplot(m_morein_tibble, aes("More abundant in", y=m_morein, fill=continents_names)) +
  geom_bar(width = 1, stat = "identity") + labs(fill='') 

more_m_pie <- more_m_bp + coord_polar("y", start=0) +
  ggtitle("Males") + theme(plot.title = element_text(hjust = 0.5, size=8)) +
  theme(axis.text.x=element_blank()) + theme(axis.title.x=element_blank()) +
  theme(axis.text.y=element_blank()) + theme(axis.title.y=element_blank()) + theme(axis.ticks.y=element_blank()) +
  geom_label(aes(label = m_morein), position = position_stack(vjust = 0.5), show.legend = FALSE)

figure_more <- ggarrange(more_f_pie, more_m_pie, ncol = 2, nrow = 1, common.legend = TRUE, legend = "right", align = "hv", font.label = list(size = 10, color = "black", face = "bold", family = NULL, position = "top"))

text <- paste(which_TE, "- How many families are most abundant in each continent?")

(figure_final <- annotate_figure(figure_more, top = text_grob(text, color = "black", size = 15, hjust = 0.55, vjust = 2), fig.lab = ""))
}
```

```{r}
#pie_chart_more_in_splitted(splitted, "class", "LINE")
#pie_chart_less_in_splitted(splitted, "class", "LINE")

#pie_chart_more_in_splitted(splitted, "class", "DNA")
#pie_chart_less_in_splitted(splitted, "class", "DNA")

#pie_chart_more_in_splitted(splitted, "class", "LTR")
#pie_chart_less_in_splitted(splitted, "class", "LTR")
```


## By population

To try to increase the resolution of our analysis, I performed  the same test but instead of calculating the mean for each Continent, I did that for each population.
```{r}
#(populations <- by_pop %>% group_by(pop, familyname, type, sex) %>% dplyr::summarise(pop_mean=mean(copynumber)) %>% arrange(familyname) %>% pivot_wider(names_from = pop, values_from = pop_mean) %>% full_join(classification, by="familyname") %>% distinct() %>% rename(TE_type = class))

#write_csv(populations, "/Volumes/Temp1/rpianezza/TE/general-analysis-by-pop/by_pop.csv")
```

The next step was performed in Python using the script "most-least_abundant_by-pop".
Here I read the resulting file.
```{r}
#regions <- HGDP %>% select(pop, country) %>% distinct()

#(leastmost_bypop <- read_csv("/Volumes/Temp1/rpianezza/TE/general-analysis-by-pop/most.csv", col_names = c("pop", "families", "sex", "class", "most_or_least")) %>% inner_join(regions, by="pop"))
```

This function was used to plot the above data using different subsets.
```{r}
#library(tidytext)

#pop_pies <- function(data, most_least, TE_type, threshold){
  
  #subset <- filter(data, most_or_least == most_least, class==TE_type) %>% mutate(pop = reorder_within(pop, families, sex)) %>% filter(families>threshold)
  
  #pops <- select(subset, pop, sex) %>% distinct() %>% select(pop) %>% separate(pop, into = c("pop", "crap")) %>% select(!(crap)) %>% pull()
  
  #ggplot(subset, aes(pop, families, fill=country)) +
    #geom_bar(stat = "identity") +
    #facet_grid(~ sex, scales = "free") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=7), axis.title.x = element_blank(), plot.title = element_text(hjust = 0.5)) + ggtitle(paste0(TE_type, " transposons: ", most_least, " abudant families by population")) + ylab("Number of repetitive sequence families")  +
    #scale_x_reordered()
  #}
```

```{r}
#pop_pies(leastmost_bypop, "most", "all", 10)
#pop_pies(leastmost_bypop, "least", "all", 10)

#pop_pies(leastmost_bypop, "most", "LINE", 3)
#pop_pies(leastmost_bypop, "least", "LINE", 1)

#pop_pies(leastmost_bypop, "most", "DNA", 10)
#pop_pies(leastmost_bypop, "least", "DNA", 7)

#pop_pies(leastmost_bypop, "most", "LTR", 3)
#pop_pies(leastmost_bypop, "least", "LTR", 5)
```