---
title: "Kruskal_Wallis test to determine the significance of various factors on the estimates of copy numbers in the HGDP dataset"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/human-data')
```
# Unbiased Population Variation of Human Transposable Elements - Script 6
This is the sixth out of eight scripts describing the creation and analysis of the dataset of human TE abundance.
This script works with the version of the HGDP dataset created in Script 2.
Here, we perform all analyses on the whole dataset as well as seperately for males and females (as we established there are differences betwene the sexes in Script 5).
```{r}
library(tidyverse)
#setwd(~/human-data)
HGDPcutoff<-read_delim("summary-HGDP/USEME_HGDP_complete_reflib6.2_mq10_batchinfo_cutoff0.01.txt",comment="#")
names(HGDPcutoff)<-c("ID","Pop","sex","Country","type","familyname","length","reads","copynumber","batch")

HGDPcutoff$ID<-as.factor(HGDPcutoff$ID)
HGDPcutoff$Pop<-as.factor(HGDPcutoff$Pop)
HGDPcutoff$sex<-as.factor(HGDPcutoff$sex)
HGDPcutoff$Country<-as.factor(HGDPcutoff$Country)
HGDPcutoff$type<-as.factor(HGDPcutoff$type)
HGDPcutoff$familyname<-as.factor(HGDPcutoff$familyname)
HGDPcutoff$copynumber<-as.numeric(HGDPcutoff$copynumber)
HGDPcutoff$reads<-as.numeric(HGDPcutoff$reads)
HGDPcutoff$batch<-as.factor(HGDPcutoff$batch)


ftecutoff<-subset(HGDPcutoff, sex=="female" & type=="te")
mtecutoff<-subset(HGDPcutoff, sex=="male" & type=="te")
```

The idea of this code is to test for the influence of specific parameters on the obtained copy number estimates.
Kruskal-Wallis is a nonparametric test, thus we do not need any assumption about the underlying distrbution of the data. 
Simply put, it is a nonparametric version of ANOVA.
We use this test to determine if different inherent properties of the data are inherently different in their corresponding copy numbers.
For example, when using the full dataset, we can confirm previous results showing a strong discrepancy in copy number estimates between males and females:
```{r}
kruskal.test(copynumber ~ sex, data = HGDPcutoff)
```
The Kurskal-Wallis test additionally allows us to explore the potential effects of other important data properties.
For the additional tests, I again look at the genders separately.
When reporting these results, it is likely that a correction for multiple testing would have  to be applied.

The most interesting test for me was to see the influence of the continent/region from which the samples were from. This test is significant with a 5% significance level for both seces, which nicely illustrates that we do actually observe regional abundance differences. Though if the significance can survive multiple testing correction I am not sure about.
```{r}
kruskal.test(copynumber ~ Country, data = mtecutoff)
kruskal.test(copynumber ~ Country, data = ftecutoff)
```

Contrarily to the Continent/Region-level, we do not observe significance when looking at variance on the population level:
```{r}
kruskal.test(copynumber ~ Pop, data = mtecutoff)
kruskal.test(copynumber ~ Pop, data = ftecutoff)
```

There is a very significant relation between copy number and length.
To think about if this is expected and if it makes sense is certainly a good idea.
```{r}
kruskal.test(copynumber ~ length, data = mtecutoff)
kruskal.test(copynumber ~ length, data = ftecutoff)
```

No significance is observed when looking at variation on the individual level, which is probably not very surprising.

```{r}
kruskal.test(copynumber ~ ID, data = mtecutoff)
kruskal.test(copynumber ~ ID, data = ftecutoff)
```

Lastly, we could look at the batch information. This information should now be meaningless regarding technical differences. However, it seems like it still has a significant effect. I believe this is solely caused by the fact that very often whole populations are in one or the other batch and it is not a random distribution.
```{r}
kruskal.test(copynumber ~ batch, data = mtecutoff)
kruskal.test(copynumber ~ batch, data = ftecutoff)
```
